### 2019-03-23 paper/sex_selection_spacing-ver2.tex
Edit first part of Results section for grammar

% [How to read tables]
Figures \ref{fig:bootstrap_low}, \ref{fig:bootstrap_med}, and  \ref{fig:bootstrap_high} 
show predicted average birth intervals, sex ratios, and probabilities of having
a birth by decade, spell, and sex composition for the three education levels separated 
by area of residence.
For each subsample, I first estimate the model described in Section 
\ref{sec:strategy} and then use the estimated coefficients to predict 
average spell length, sex ratio, and the probability of having a birth in that spell.

The expected average duration is calculated as follows.
For each woman in a sample, I calculate her probability of giving birth for each period
in the spell and use these probabilities as weights to calculated her expected duration 
for that spell.
I then take the average of these expected durations using her probability of having a
birth by the end of the spell as weights.

The predicted sex ratio is the weighted average over the women in the sample's 
individual predicted sex ratios, using the individual probabilities of having
given birth by the end of the spell as weights.
Each woman's predicted sex ratio is the weighted average of the predicted
percentage boys over each period in the spell calculated using equation 
(\ref{eq:probability_son}) and the probability of giving birth in each period 
as weights.%
\footnote{
Imagine a spell has two periods and that the estimated percentage boys for a
woman are 54\% and 66\% and that the likelihood of having a birth is 20\% and 40\%.
The likelihood of having a birth is the change in the survival curve; 
in other words, there is 40\% chance that she will not have given birth by the end of 
this spell.
This woman's percentage boys is then $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$.
}
The predicted sex ratio captures the percent boys that will have been born to women 
in the sample when childbearing for that spell is over.

For legibility none of the graphs show standard errors.
Instead, the standard errors, together with the graphed values, are shown in Appendix 
Tables TK, TK, and TK.
The standard errors are based on bootstrapping for all three measures,
where the model is repeatedly estimated using resampling with replacement.

@@

Each predicted percent boys is tested against the natural percentage 
boys using the bootstrapped standard errors.
The natural sex ratio is approximately 105 boys to 100 girls or
51.2\% \citep{ben-porath76b,jacobsen99,Portner2015b}.
The predicted percentage boys may differ from the natural rate because of 
natural variation, any remaining recall error not corrected for, or 
sex selection.

@@

The previous literature finds that women with no education are the least likely to use 
sex selection, which suggests that they would instead follow a traditional strategy to son 
preference with higher fertility and shorter birth intervals the fewer sons they have.
This pattern is supported by Figure \ref{fig:bootstrap_low}, but with important
differences across spells.
Most of the women without education live in rural areas, and I therefore focus on rural 
women.%
\footnote{
Only between ten and twenty percent of the surveyed women without education live in urban 
areas, and this proportion has been decreasing over time.
}

The initial high fertility of women with no education shows clearly in the results.
In the 1972--1984 period almost all women had at least four children, irrespectively
of the sex composition of the children.
Over time, however, the likelihood of having a third or fourth child decreased for women 
with at least one son, and more so the more sons they had.
Hence, fertility have begun to fall even among women with no education, as long as they 
have at least one son.
Two additional fertility trends stand out.
First, essentially all women still have a second birth, whether the first child was
a boy or a girl.
Second, for women with only girls there is almost no reduction in the probability of 
having a birth across time and across spells.

Birth intervals have increased in length over time, although the increases are not
uniform over spells and sex compositions of prior children.
The smallest increases have been for the second spell, with most increases between one
and two months.
Interestingly, son preference appear to have little effect on spacing for the second
spell, with very small, albeit statistically significant, differences in birth intervals 
depending on the sex of the first child.
This small difference is in line with the high likelihood of having a second child.
The second spell also shows convergence in birth intervals, especially in the urban
areas where the predicted average duration is now the same whether the first child was
a boy or a girl.

Where son preference in birth intervals shows most clearly is for the fourth spell.
Despite the increase in spacing even for those with three girls, the increases in 
spacing have been substantially larger for families with at least one son.
As for fertility, the differences are larger the more sons a family has.

Despite the general finding that use of sex selection is confined to higher education
women, the predicted sex ratios for those with only girls is statistically significantly
higher than the natural ratio for the third and fourth spell.
Hence, it is possible that at least part of the increased  birth intervals for those
with only girls comes from the use of sex selection and that the differences in
birth spacing across sex compositions would be even larger in the absense of sex selection.

@@

Figure \ref{fig:bootstrap_med} shows that women with one to seven years of education 
follow a fertility pattern broadly similar to those with no education.
As expected, fertility is lower than for women with no education, but still show
substantial son preference in the parity progression behavior.
Except for the second spell, the likelihood of having a birth by the end of spell is
substantially lower if there has been at least one male birth than if there are only girls.
Despite this, the probability of a birth in the third and fourth spells has also been 
falling for those with only girls.
Having a second birth is still close to universal, although in urban areas it is down to
just over 85 percent for those with a son as their first child.

The spacing between births has increased over the four decades, and the increases are 
larger than for women without any education across all spells.
Contrary to women without education, however, there has not been an increased divergence 
in birth intervals lengths across sex compositions for the third and fourth spells.
On the contrary, for the third spell the largest increases in birth intervals are for women with
two daughters and no sons, with an increase of about seven months across the four decades.
The results is that for urban women the standard spacing pattern reversed, so the longest 
predicted spacing is now for those with only girls, while for rural women the predicted 
average birth intervals are now almost identical across sex compositions of prior children.
For the fourth spell, the increases in spacing for women with only girls have kept pace
with the other sex compositions in rural areas.%
\footnote{
Because of low number of observations the pattern for the fourth spell is more noisy for 
urban women.
}

The substantial increases in spacing for women with only girls is not evidence of lower
son preference as the changes in the predicted sex ratios show.
For both the third and fourth spell the predicted sex ratio in both urban and rural
areas is statistically significantly above the natural sex ratio at between 55.5 and 57.7 
percent boys.
The elevated sex ratios strongly suggests that sex selection is driving the increases in
birth intervals for women with only girls, and that without sex selection we would 
instead have observed a diverging pattern in birth intervals across sex composition as
for women without education.

@@

Women with eight or more years of education are those expected to have the lowest 
fertility and the highest use of sex selection based on the prior literature.
Figure \ref{fig:bootstrap_high} shows that not only do this group, indeed, have lower 
fertility than the two prior groups, it is also falling rapidly over time.
Of women with one birth, the probability of having a second birth when the first-born
is a son is only around 70 percent for urban women and around 80 percent for rural women,
and for those with a first-born daughter it is around 80 percent for urban women and 90 
percent for rural women.

The rapid decline in fertility for the most educated women is even more clear for the 
third and fourth spells when the woman has given birth to at least one son.
At the beginning of the sample, urban women with at least one boy had a close to 75
percent likelihood of having a birth, but that has fallen to only around 25 percent.
However, son preference still shows clearly in the parity progression probability for 
women without a son, with the decline for the third and fourth spell substantially
slower for both urban and rural women without a son than for those with at least one son.
The falling fertility means that relatively few women makes it to the fourth spell,
which, therefore, has more noise than the other spells.

Average birth spacing for the second spell is almost identically across sex compositions 
for both urban and rural women, and increased by around five months for rural women and
by almost ten months for urban women over the four decades.
These similarities hides, however, what would have been a standard son preference spacing 
pattern with significantly shorter spacing after a first-born girl than after a first-born 
boy if there had not been a substantial amount of sex selection.
For both of the last two periods the predicted sex ratio for women with a daughter as 
their first-born child is above 55 percent for both urban and rural areas.

The tremendous impact that sex selection can have on birth spacing is illustrated 
particularly vividly by the changes over time for the third spell.
In the 1972-1984 period the predicted average birth intervals for the three possible sex 
compositions were less than two months apart.
For the later periods, however, the highest predicted average spacing was for women
with only daughters with most of the differences to the other sex compositions 
statistically significant.
In urban areas the predicted average birth interval for a woman with only daughters 
increased by almost a full year from 26.8 to 38.3 months.
What is more, this increase is likely an underestimate.
The sex ratios for the 1972-84 period are substantially higher than the natural sex ratio 
despite sex selection not being widely available until around 1985.
If this early elevated sex ratio is because of recall error it means that the actual birth 
intervals would have been substantially shorter than the ones predicted here and the
increase over time larger.

The reversal pattern also shows for the fourth spell for urban women, although not for
rural women.
An important caveat is that because of the falling fertility relatively few women make
it to the fourth spell, and even if they do we do not observe a birth by the time of the
survey.
Of the 11,886 women across urban and rural areas observed for the fourth spell for the 
2005-2016 period more than 70 percent did not have a birth at the time of the survey.
Even though the censoring is less prevalent for the earlier periods this is countered by
those samples being substantially smaller.
Despite this, the pattern for the fourth spell is similar to that of the third spell.

The predicted sex ratios at the end of the spells show that this reversal in spacing 
patterns is not the result of a declining son preference but instead corresponds to a 
substantially more male-biased sex ratio as use of sex selection spread.
For urban women, the predicted sex ratio with only girls is consistently above 60\% boys 
for the third and fourth spells.
For rural women, the predicted sex ratios are lower than for urban women but still 
substantial and statistically significant at close to 60\%.

>>>>

% [How to read tables]
Figures \ref{fig:bootstrap_low}, \ref{fig:bootstrap_med}, and  \ref{fig:bootstrap_high} 
show predicted average birth intervals, sex ratios, and probabilities of having
a birth by decade, spell, and sex composition for the three education levels separated 
by the area of residence.
For each subsample, I first estimate the model described in Section 
\ref{sec:strategy} and then use the estimated coefficients to predict 
average spell length, sex ratio, and the probability of having a birth in that spell.

I calculate the expected average duration as follows.
For each woman in a sample, I calculate her probability of giving birth for each period
in the spell and use these probabilities as weights to calculated her expected duration 
for that spell.
I then take the average of these expected durations using her probability of having a
birth by the end of the spell as weights.

The predicted sex ratio is the weighted average over the women in the sample's 
individual predicted sex ratios, using the individual probabilities of having
given birth by the end of the spell as weights.
Each woman's predicted sex ratio is the weighted average of the predicted
percentage of boys over each period in the spell calculated using equation 
(\ref{eq:probability_son}) and the probability of giving birth in each period 
as weights.%
\footnote{
Imagine a spell has two periods and that the estimated percentages of boys for a
woman are 54\% and 66\% and that the likelihood of having a birth is 20\% and 40\%.
The likelihood of having a birth is the change in the survival curve; 
in other words, there is a 40\% chance that she will not have given birth by the end of 
this spell.
This woman's percentage of boys is then $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$.
}
The predicted sex ratio captures the percent boys that will have been born to women 
in the sample when childbearing for that spell is over.

For legibility, none of the graphs show standard errors.
Instead, the standard errors, together with the graphed values, are shown in Appendix 
Tables 
\ref{tab:bootstrap_duration_avg_sex_ratio_low_all}, 
\ref{tab:bootstrap_duration_avg_sex_ratio_med_all}, and 
\ref{tab:bootstrap_duration_avg_sex_ratio_high_all}.
The standard errors are based on bootstrapping for all three measures,
where the model is repeatedly estimated using resampling with replacement.

@@

Each predicted percent of boys is tested against the natural percentage of
boys using the bootstrapped standard errors.
The natural sex ratio is approximately 105 boys to 100 girls or
51.2\% \citep{ben-porath76b,jacobsen99,Portner2015b}.
The predicted percentage boys may differ from the natural rate because of 
natural variation, any remaining recall error not corrected for, or 
sex selection.

@@

The previous literature finds that women with no education are the least likely to use 
sex selection, which suggests that they would instead follow a traditional strategy to son 
preference with higher fertility and shorter birth intervals the fewer sons they have.
This pattern is supported by Figure \ref{fig:bootstrap_low}, but with significant
differences across spells.
Most of the women without education live in rural areas, and I, therefore, focus on rural 
women.%
\footnote{
Only between ten and twenty percent of the surveyed women without education live in urban 
areas, and this proportion has been decreasing over time.
}

The initial high fertility of women with no education shows clearly in the results.
In the 1972--1984 period almost all women had at least four children, irrespectively
of the sex composition of the children.
Over time, however, the likelihood of having a third or fourth child decreased for women 
with at least one son, and more so the more sons they had.
Hence, fertility has begun to fall even among women with no education, as long as they 
have at least one son.
Two additional fertility trends stand out.
First, virtually all women still have a second birth, whether the first child was
a boy or a girl.
Second, for women with only girls, there is almost no reduction in the probability of 
having a birth across time and spells.

Birth intervals have increased in length over time, although the increases are not
uniform over spells and sex compositions of prior children.
The smallest increases have been for the second spell, with most increases between one
and two months.
Interestingly, son preference appears to have little effect on spacing for the second
spell, with very small, albeit statistically significant, differences in birth intervals 
depending on the sex of the first child.
This small difference is in line with the high likelihood of having a second child.
The second spell also shows convergence in birth intervals, especially in the urban
areas where the predicted average duration is now the same whether the first child was
a boy or a girl.

Where son preference in birth intervals shows most clearly is for the fourth spell.
Despite the increase in spacing even for those with three girls, the increases in 
spacing have been substantially stronger for families with at least one son.
As for fertility, the more sons a family has, the more significant are the differences.

Despite the prior literature finding that women with higher education are the primary user 
of sex selection, the predicted sex ratios for those with only girls are statistically 
significantly higher than the natural ratio for the third and fourth spell.
Hence, it is possible that at least part of the increased  birth intervals for those
with only girls comes from the use of sex selection and that the differences in birth 
spacing across sex compositions would be even more substantial in the absence of sex 
selection.

@@

Figure \ref{fig:bootstrap_med} shows that women with one to seven years of education 
follow a fertility pattern broadly similar to those with no education.
As expected, fertility is lower than for women with no education, but still a
strong son preference in the parity progression behavior.
Except for the second spell, the likelihood of having a birth by the end of the spell is
substantially lower if there has been at least one male birth than if there are only girls.
Despite this, the probability of a birth in the third and fourth spells has also been 
falling for those with only girls.
Having a second birth is still close to universal, although in urban areas it is down to
just over 85 percent for those with a son as their first child.

The spacing between births has increased over the four decades, and the increases are 
larger than for women without any education across all spells.
Contrary to women without education, however, there has not been an increased divergence 
in birth intervals lengths across sex compositions for the third and fourth spells.
On the contrary, for the third spell, the most substantial increases in birth intervals are for 
women with two daughters and no sons, with an increase of about seven months across the 
four decades.
The result is that for urban women the standard spacing pattern reversed, so the longest 
predicted spacing is now for those with only girls, while for rural women the predicted 
average birth intervals are now almost identical across sex compositions of prior children.
For the fourth spell, the increases in spacing for women with only girls have kept pace
with the other sex compositions in rural areas.%
\footnote{
Because of the low number of observations, the pattern for the fourth spell is noisier 
for urban women.
}

The substantial increases in spacing for women with only girls is not evidence of lower
son preference as the changes in the predicted sex ratios show.
For both the third and fourth spell the predicted sex ratio in both urban and rural
areas is statistically significantly above the natural sex ratio at between 55.5 and 57.7 
percent boys.
The elevated sex ratios strongly suggest that sex selection is driving the increases in
birth intervals for women with only girls and that without sex selection we would 
instead have observed a diverging pattern in birth intervals across sex composition as
for women without education.

@@

Women with eight or more years of education are those expected to have the lowest 
fertility and the highest use of sex selection based on the prior literature.
Figure \ref{fig:bootstrap_high} shows that this group not only have lower fertility than 
the two previous groups; fertility is also falling rapidly over time.
Of women with one birth, the probability of having a second birth when the first-born
is a son is only around 70 percent for urban women and around 80 percent for rural women,
and for those with a first-born daughter, it is around 80 percent for urban women and 90 
percent for rural women.

The rapid decline in fertility for the most educated women is even more clear for the 
third and fourth spells when the woman has given birth to at least one son.
At the beginning of the sample, urban women with at least one boy had a close to 75
percent likelihood of having a birth, but that has fallen to only around 25 percent.
However, son preference still shows clearly in the parity progression probability for 
women without a son, with the decline for the third and fourth spell substantially
slower for both urban and rural women without a son than for those with at least one son.
The falling fertility means that relatively few women make it to the fourth spell,
which, therefore, has more noise than the other spells.

Average birth spacing for the second spell is almost identically across sex compositions 
for both urban and rural women and increased by around five months for rural women and
by almost ten months for urban women over the four decades.
These similarities hide, however, what would have been a standard son preference spacing 
pattern with significantly shorter spacing after a first-born girl than after a first-born 
boy if there had not been a substantial amount of sex selection.
For both of the last two periods, the predicted sex ratio for women with a daughter as 
their first-born child is above 55 percent for both urban and rural areas.

The tremendous impact that sex selection can have on birth spacing is illustrated 
particularly vividly by the changes over time for the third spell.
In the 1972-1984 period, the predicted average birth intervals for the three possible sex 
compositions were less than two months apart.
For the later periods, however, the highest predicted average spacing was for women
with only daughters with most of the differences to the other sex compositions 
statistically significant.
In urban areas, the predicted average birth interval for a woman with only daughters 
increased by almost a full year from 26.8 to 38.3 months.
What is more, this increase is likely an underestimate.
The sex ratios for the 1972-84 period are substantially higher than the natural sex ratio 
despite sex selection not being widely available until around 1985.
If this early, elevated sex ratio is because of recall error, it means that the actual birth 
intervals would have been substantially shorter than the ones predicted here and the
increase over time larger.

The reversal pattern also shows for the fourth spell for urban women, although not for
rural women.
An important caveat is that because of the falling fertility relatively few women make
it to the fourth spell, and even if they do, we do not observe a birth by the time of the
survey.
Of the 11,886 women across urban and rural areas observed for the fourth spell for the 
2005-2016 period more than 70 percent did not have a birth at the time of the survey.
Even though the censoring is less prevalent for the earlier periods, this is countered by
those samples being substantially smaller.
Despite this, the pattern for the fourth spell is similar to that of the third spell.

The predicted sex ratios at the end of the spells show that this reversal in spacing 
patterns is not the result of a declining son preference but instead corresponds to a 
substantially more male-biased sex ratio as the use of sex selection spread.
For urban women, the predicted sex ratio with only girls is consistently above 60\% boys 
for the third and fourth spells.
For rural women, the predicted sex ratios are lower than for urban women but still 
substantial and statistically significant at close to 60\%.
